### 2016-02-13 Paper_v4.tex
Figs 4 and 5

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Dynamic Optimization with Particle Swarms (DOPS): A meta-heuristic for parameter estimation in biochemical models}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Christine Shoemaker$^{\dag}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {$^{\dag}$School of Civil and Environmental Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Parameter estimation in biochemical models}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{PLoS ONE}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Mathematical modeling is a powerful tool to analyze, and ultimately design biochemical networks.
However, the estimation of biochemical model parameters is a significant challenge.
Parameter estimation in biochemical models typically involves expensive function evaluations and noisy data, making it difficult to quickly obtain optimal solutions.
Biochemical models often also have many local extrema which further complicates parameter estimation.
Toward these challenges, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel global meta-heuristic that combined features of multi-swarm particle swarm optimization with dynamically dimensioned search (DDS).
DOPS uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors, the best of which is greedily updated using dynamically dimensioned search.
We tested the performance of DOPS on a model of human coagulation cascade.
We performed $\mathcal{T}$ = 25 trials with $\mathcal{N}$ = 4000 function evaluations per trial, and compared the performance of DOPS with other commonly
used meta-heuristics such as differential evolution (DE), simulated annealing (SA) and dynamically dimensioned search (DDS).
We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between simulations and experimental measurements.
Lastly, we tested the performance of DOPS on commonly used test functions for global optimization and on published biochemical parameter estimation benchmark problems.
For the wide range of problems that we considered, DOPS outperformed other meta-heuristic approaches despite a limited number of function evaluations.
Taken together, DOPS is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.

\vspace{0.1in}
{\noindent \textbf{Keywords:}~Parameter identification, Meta-heuristic optimization, Biochemical modeling}

@@

Cells process nutrients and respond to changes in their environment using complex biochemical networks. These networks contain thousands of components
interconnected through nonlinear enzyme catalyzed reactions. Mathematical modeling has evolved as a powerful paradigm to analyze, and ultimately design these complex networks \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}. Mathematical modeling of biochemical networks is often an iterative process.
First, models are formulated from biochemical knowledge, and then model parameters are estimated using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
Parameter estimation is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and model simulations is minimized using an optimization strategy \cite{moles2003parameter}. Optimal parameters obtained from model training are then used to validate the model on unseen experimental data.
If validation fails, model construction and calibration are repeated iteratively until satisfactory results are obtained.

Parameter estimation is a major challenge in the iterative development of biochemical models.
Although parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a unique estimation challenge.
Often optimization problems involving biological systems are non-linear and multi-modal i.e. typical models often have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}.
Non-linearity coupled with multi-modality generally renders local optimization techniques such as pattern search \cite{hooke1961direct}, Nelder-Mead simplex methods \cite{nelder1965simplex},
steepest descent methods or Levenberg-Marquardt \cite{more1978levenberg} incapable of reliably obtaining optimal solutions as they generally terminate at local minimum.
Though deterministic global optimization techniques (for example algorithms based on branch and bound) can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global},
the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of knowledge about the objective function hampers these techniques.

Meta-heuristic stochastic optimization approaches like Genetic Algorithms (GAs) [ADI-REF], Simulated Annealing (SA) \cite{kirkpatrick1983optimization},
Evolutionary Programming [ADI-REF] and population based searches like Differential Evolution (DE) \cite{storn1997differential} have shown promise on nonlinear multi-modal problems \cite{sun2012parameter}.
These techniques do not make any assumptions about the structure of objective function nor do they require \textit{a~priori} information about the objective function.
Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.
Mendes et al. \cite{mendes1998non} used simulated annealing to estimate rate constants for the irreversible inhibition of HIV proteinase,
Modchang et al. \cite{modchang2008mathematical} used a genetic algorithm to estimate parameters for a model of signal transduction [ADI-EXPAND],
differential evolution approaches have also been effective on various biological problems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}.
Tashkova et al. compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective \cite{tashkova2011parameter}.
Banga and co-workers have also successfully applied scatter-search methods to estimate parameters on non-linear biological processes \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter}.
Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near globally optimal solution obtained using a meta-heuristic is refined using a local search have also become popular.
Villaverde et al. combined scatter search with local search methods for parameter estimation in a collection of systems biology models \cite{villaverde2015biopredyn}.
Fan et al. recently showed that population based meta-heuristics along with decomposition based methods can be also used to model gene circuits from mRNA data \cite{fan2015parameter}.
Despite these successes, a major drawback of most metaheuristic approaches is the large number of function evaluations required to explore the parameter space.
Typically as models grow in size and complexity, evaluation of the objective function becomes computationally expensive.
Thus performing a large number of function evaluations is not desirable (and perhaps not feasible).

%In many of these high dimensional problems approaching an exact solution may not be necessary. Gutenkust et al. \cite{gutenkunst2007universally} showed that a number of systems biology models are 'sloppy'. Sloppy systems have specific parameter combinations that largely define the dynamics of the system. Large perturbations to the rest of the parameters does not greatly impact the system dynamics. Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}. Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions.

%Because the neighborhood in DDS is the entire parameter space, DDS is a global optimization technique although it is greedy.

In this study, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel meta-heuristic that combines the global search capability of multi-swarm particle swarm optimization and dynamically dimensioned search (DDS). The objective of DOPS is to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
DOPS uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamically dimensioned search. We first considered a model of human coagulation cascade to test the performance of DOPS. Coagulation is a large, complex biochemical network involving strong positive feedback.
We then tested the performance of DOPS on commonly used test functions for global optimization (Ackley and Rosenbrock) and published biochemical parameter estimation benchmark problems \cite{villaverde2015biopredyn}. DOPS outperformed common meta-heuristic approaches like Differential Evolution (DE), Simulated Annealing (SA) and dynamically dimensioned search (DDS) on both the test functions and the coagulation model.
DOPS also performed well on the benchmark problems where it outperformed enhanced scatter search (eSS) and recovered the nominal parameters with only 4000 function evaluations ([ADI-compared with what?])
across all the benchmark problems considered.
Taken together, these studies suggest DOPS is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.

@@

\section*{Results}

\subsection*{Optimization problem formulation.}
The problem of parameter estimation in a dynamic biological model consists of finding an optimal parameter vector
which minimizes the difference between model simulations and $\mathcal{E}$ experimental measurements. This difference is quantified by an objective function $K\left(\mathbf{p}\right)$
which is typically the Euclidean norm of the simulation error subject to problem specific and parameter bounds constraints:

\begin{equation}\label{eq:static_opt}
\begin{aligned}
& \text{minimize}
& & K(\mathbf{p})=\sum_{i=1}^\mathcal{E} \left(g_i(t_i,\mathbf{x,p,u})-y_i\right)^2 \\
& \text{subject to}
&&\dot{\mathbf{x}}=\mathbf{f}(t,\mathbf{x}(t,\mathbf{p}),\mathbf{u}(t),\mathbf{p})\\
&&&\mathbf{x}(t_0) = \mathbf{x}_0\\
&&&\mathbf{c}(t,\mathbf{x,p,u}) \geqslant \mathbf{0} \\
&&& \mathbf{p}^L \leqslant \mathbf{p} \leqslant \mathbf{p}^U\\
\end{aligned}
\end{equation}
where t is time, $\mathbf{x}\left(t,\mathbf{p}\right)$ is the state variable vector with an initial state $\mathbf{x}_{0}$, $\mathbf{u}(t)$ is a model input vector,
$\mathbf{f}$ is the system of model equations (e.g., differential equations or algebraic constraints) and $\mathbf{p}$ is the model parameter vector.
The parameter search (or model simulations) can be subject to $\mathbf{c}$ linear or non-linear constraints, and parameter bound constraints where
p\textsuperscript{L} and p\textsuperscript{U} denote the lower and upper parameter bounds, respectively.
The problem eventually is to find:
\begin{equation}\label{eq:static_opt_2}
\mathbf{p^{*}}= \arg\min_{\mathbf{p}} K\left(\mathbf{\mathbf{p}}\right)
\end{equation}

\subsection*{Dynamic optimization with particle swarms (DOPS).}
DOPS is a novel meta-heuristic which combines multi-swarm particle swarm methods with dynamically dimensioned search (Fig. \ref{fig-algorithm}).
The goal of DOPS is to estimate optimal or near optimal parameter vectors for high-dimensional biological models within a specified number of function evaluations.
Toward this objective, DOPS begins by using a particle swarm search and then dynamically switches, using an adaptive switching criteria, to the DDS search phase.
We began the particle swarm phase by randomly initializing a swarm of $\mathcal{K}$-dimensional particles (represented as $z_{i}$), wherein each particle corresponded to a $\mathcal{K}$-dimensional parameter vector.
After initialization, particles were randomly partitioned into $k$ equal sized sub-swarms $\mathcal{S}_{1},\hdots,\mathcal{S}_{k}$.
Thereafter within each sub-swarm $\mathcal{S}_{k}$, particles were updated according to the rule:
\begin{equation}\label{eqn:update-rule}
	\mathbf{z}_{i,j} = \theta_{1,j-1}\mathbf{z}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{z}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{G}_{k} - \mathbf{z}_{i,j-1}\right)
\end{equation}
where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ were adjustable parameters, $\mathcal{L}_{i}$ denotes the best solution found by particle $i$ within sub-swarm $k$ for function evaluation $1\rightarrow j-1$, and
$\mathcal{G}_{k}$ denotes the best solution found over all particles within sub-swarm $\mathcal{S}_{k}$.
The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model parameters ($\mathcal{K}\times{1}$).
Equation (3) is similar to the general particle swarm update rule, however, it does not contain velocity terms.
In DOPS the parameter $\theta_{1,j-1}$ depends upon the function evaluations and is updated according to:
\begin{eqnarray}
	\mathbf \theta_{1,j}&=&\frac{(\mathcal{N}-{j})*(\mathbf{w}_{max}-\mathbf{w}_{min}))}{(\mathcal{N}-{1})} + \mathbf{w}_{min}
\end{eqnarray}
where $\mathcal{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights, respectively.
While updating the particles, we made sure all dimensions of the solution represented by the particle were within bounds using a set of reflection boundary conditions:

\begin{eqnarray*}
\end{eqnarray*}

\begin{algorithm}
\If{$z_{i,j}^{old} < z_{i}^{min}$}{
  ${z}_{i,j}^{new}={z}_{i,j}^{old}+(z_{i}^{min} - {z}_{i,j}^{old})$
  \If{${z}_{i,j}^{new}>z_{i}^{max}$}{
    ${z}_{i,j}^{new}=z_{i}^{max}$
  }
}
\If{$z_{i,j}^{old} > z_{i}^{max}$}{
  ${z}_{i,j}^{new}={z}_{i,j}^{old}+({z}_{i,j}^{old}-z_{i}^{max})$
  \If{${z}_{i,j}^{new}<z_{i}^{min}$}{
    ${z}_{i,j}^{new}=z_{i}^{min}$
  }
}
\end{algorithm}

After every $\mathcal{M}$ function evaluations, particles were randomly redistributed to a new sub-swarm, and updated according to Eqn. \eqref{eqn:update-rule}.
This process continued till $\mathcal{F}*\mathbf{N}$ number of functions evaluations, where $\mathcal{F}$ is the fraction of evaluations with the multi-swarms.
The fraction of evaluations within the swam phase is based on the drop in error.
If the error does not drop by 1\% for a certain number of evaluations the search within swarm phase is terminated.
At the end of these function evaluations, we froze all the solutions represented by various particles and chose the particle with best solution among $\mathcal{G}_{1} \cdots \mathcal{G}_{NS}$
as the initial candidate vector $\mathcal{G}$ for the remaining $({1}-\mathcal{F})*\mathbf{N}$ number of function evaluations.
This particle was then updated according to the rule:
\begin{equation}
  \mathcal{G}_{new}(\mathbf{J})=\begin{cases}
    \mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J}), & \text{if $\mathcal{G}_{new}(\mathbf{J})<\mathcal{G}(\mathbf{J})$}.\\
    \mathcal{G}(\mathbf{J}), & \text{otherwise}.
  \end{cases}
\end{equation}
where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$,
and $\sigma$ is the amplitude of perturbation:
\begin{equation}
	\sigma = \mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
\end{equation}
where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension. The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.  This function is a monotonically decreasing function that decreases with the number of function evaluations. $\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function:
\begin{eqnarray}
	%\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
	\mathcal{P}_{j}&=&{1}-\log(j/ (({1}-\mathcal{FR})*\mathbf{N}))
\end{eqnarray}
%where $\beta$ is the perturbation frequency probability modulator.
Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase. These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$. The decrease of weight function in equation 4, reflection boundary conditions in equation 5, the update function described in equations 6 and 7, the selection probability in equation 8 are the means by which DDS ideas are incorporated into DOPS.
%Equation 5 describing the reflection boundary conditions, equations 6,7 and 8 are the means by which DDS ideas are incorporated into DOPS.
The fraction of evaluations $\mathcal{FR}$  within the swarm phase is based on a switching strategy wherein the switch from swarm phase to DDS phase happens when the error due to the best solution does not drop more than 1\% of the original error, continuously for more than a prescribed number of function evaluations. This allows the solution to quickly jump out of local optima and avoid any convergence issues that are generally associated with swarm based searches.

>>>>

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Dynamic Optimization with Particle Swarms (DOPS): A meta-heuristic for parameter estimation in biochemical models}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Christine Shoemaker$^{\dag}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {$^{\dag}$School of Civil and Environmental Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Parameter estimation in biochemical models}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{PLoS ONE}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Mathematical modeling is a powerful tool to analyze, and ultimately design biochemical networks. However, estimation of biochemical model parameters is a significant challenge. Parameter estimation in biochemical models generally involves expensive function evaluations, making it difficult to quickly obtain optimal solutions. Additionally, biochemical models often have many local minima or maxima which further complicates the parameter estimation problem. In this study, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel meta-heuristic that combined features of multi-swarm particle swarm optimization with dynamically dimensioned search (DDS). DOPS uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors of which the best one is greedily updated using dynamically dimensioned search. We first tested the performance of DOPS on a model of human coagulation cascade. We performed 25 trials with 4000 function evaluations per trial, and compared the performance of DOPS with other commonly
used meta-heuristic approaches such as differential evolution (DE), simulated annealing (SA) and dynamically dimensioned search (DDS). We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between model simulations and experimental measurements. Thereafter we tested the performance of DOPS on commonly used test functions for global optimization (Ackley and Rosenbrock) and on published biochemical parameter estimation benchmark problems.  For the wide range of problems that we considered, DOPS outperformed other meta-heuristic approaches despite a limited number of function evaluations. Taken together, DOPS is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.

{\noindent \textbf{Keywords:}~Parameter identification, Meta-heuristic optimization, Biochemical modeling}

@@

Cells process nutrients and respond to changes in their environment using complex biochemical networks. These networks contain thousands of components
interconnected through nonlinear enzyme catalyzed reactions. Mathematical modeling has evolved as a powerful paradigm to analyze, and ultimately design these complex networks \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}. Mathematical modeling of biochemical networks is often an iterative process.
First, models are formulated from biochemical knowledge, and then model parameters are estimated using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
Parameter estimation is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and model simulations is minimized using an optimization strategy \cite{moles2003parameter}. The optimal parameters obtained from training are then used to validate the model on hitherto unseen experimental data. If validation fails, model construction and calibration are repeated iteratively until satisfactory results are obtained.

Parameter estimation is a major challenge in the development of large biochemical models.
Although parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a unique estimation challenge.
Most of the optimization problems pertaining to biological systems are non-linear and multi-modal i.e. most of them have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}. The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex}, steepest descent method or Levenberg-Marquardt method \cite{more1978levenberg} incapable of reliably obtaining an optimal solution since they generally stop at the first local minimum. Though deterministic global optimization techniques (for example algorithms based on branch and bound framework) can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.

Meta-heuristic stochastic optimization approaches like Genetic Algorithms (GAs), Simulated Annealing (SA) \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution (DE) \cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}. They do not make any assumptions about the structure of objective function or require any \textit{a priori} information about the objective function. Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.  Mendes et al. \cite{mendes1998non} used Simulated Annealing to estimate rate constants for the irreversible inhibition of HIV proteinase, Modchang et al. \cite {modchang2008mathematical} used Genetic Algorithms to estimate parameters for a mathematical model of signal transduction, Differential Evolution based approaches have been an effective approach on various systems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}. Tashkova et al. \cite{tashkova2011parameter} compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective. Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter} to estimate parameters on non-linear biological processes. Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near globally optimal solution that is obtained with a meta-heuristic is further refined using a local search have also become popular. Villaverde et al. \cite{villaverde2015biopredyn} combined scatter search with local search methods for parameter estimation in large scale systems biology models. Fan et al. recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}. Despite these successes, a  major drawback with most of these metaheuristic approaches is the vast number of objective function evaluations they take in arriving at good search regions. As the models grow in size and complexity and the number of dimensions of parameter vector increases, evaluation of the objective function becomes computationally expensive. Thus performing a large number of evaluations is not computationally feasible.

In many of these high dimensional problems approaching an exact solution may not be necessary. Gutenkust et al. \cite{gutenkunst2007universally} showed that a number of systems biology models are 'sloppy'. Sloppy systems have specific parameter combinations that largely define the dynamics of the system. Large perturbations to the rest of the parameters does not greatly impact the system dynamics. Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}. Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions.

In this study, we developed Dynamic Optimization with Particle Swarms (DOPS), a novel meta-heuristic that combines the global search capability of multi-swarm particle swarm optimization and dynamically dimensioned search (DDS). The objective of DOPS is to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations. DOPS uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamically dimensioned search. Because the neighborhood in DDS is the entire space, DDS is a global optimization technique although it is greedy. We first considered a model of human  coagulation cascade to test the performance of DOPS. Coagulation is a large, complex biochemical network that is tightly regulated with several feedback loops.  We then tested the performance of DOPS on commonly used test functions for global optimization (Ackley and Rosenbrock) and published biochemical parameter estimation benchmark problems \cite{villaverde2015biopredyn}. DOPS outperformed common meta-heuristic approaches like Differential Evolution (DE), Simulated Annealing (SA) and dynamically dimensioned search (DDS) on the test functions and the coagulation model. It also performed very well on the benchmark problems where it outperformed enhanced scatter search (eSS) and recovered the nominal parameters with only 4000 function evaluations across all the benchmark problems considered.

@@

\section*{Problem Formulation}

The problem of parameter estimation in dynamic biological systems consists of finding an optimal parameter vector that results in the best fit between the model predictions and observed experimental data.The cost function or objective that measures the goodness of fit is generally the Euclidean norm which is minimized under certain constraints. This can be mathematically expressed as follows.

\begin{equation}\label{eq:static_opt}
\begin{aligned}
& \text{minimize}
& & K(\mathbf{p})=(\sum_{i=1}^N (g_i(t_i,\mathbf{x,p,u})-y_i)^2) \\
& \text{subject to}
&&\dot{x}=f(t,\mathbf{x(t,\mathbf{p}),p,u(\mathbf{t})})\\
&&&x(t_0) = x_0\\
&&&c(t,\mathbf{x,p,u}) \geqslant 0 \\
&&& \mathbf{p}^L \leqslant \mathbf{p} \leqslant \mathbf{p}^U\\
\end{aligned}
\end{equation}

where t is time, x is the state variable with an initial state x\textsubscript{0}, u is the input vector \, p is the parameter vector and f is the system of equations (like differential equations or set of algebraic constraints), c in the set of linear and non-linear constraints, g is the set of observables and p\textsuperscript{L} and p\textsuperscript{U} are the lower and upper bounds. Thus the problem eventually is to find {p}\textsuperscript{*} where

\begin{equation}\label{eq:static_opt_2}
\mathbf{p^{*}}= \arg\min_{p} K(\mathbf{p})
\end{equation}

\section*{Results}
\subsection*{Dynamic Optimization with Particle Swarms (DOPS).}
DOPS is a novel meta-heuristic that combines multi-swarm based particle swarm methods with dynamically dimensioned search (Fig. \ref{fig-algorithm}).
The goal of this approach is to obtain optimal or near optimal parameters for high-dimensional complex biological systems within a pre-specified number of function evaluations. We randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to a $\mathcal{K}$-dimensional parameter vector. After initialization, the particles were grouped into different sub-swarms randomly. Thereafter within each sub-swarm ${S}_{k}$,  particles were updated according to the following rule.
\begin{eqnarray}
	\mathbf{x}_{i,j} &=&\theta_{1,j-1}\mathbf{x}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{x}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{GL}_{k} - \mathbf{x}_{i,j-1}\right)
\end{eqnarray}
where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best solution found by particle $i$ within the sub-swarm till function evaluation $j$-$1$, and
$\mathcal{GL}_{k}$ denotes the best solution found over the population of all particles within the sub-swarm ${S}_{k}$. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). Equation (3) is suggested by the general form of particle swarm equations. However this equation does not contain the velocity terms that are generally associated with particle swarm optimization. In our algorithm the parameter $\theta_{1,j-1}$ depends on the function evaluations and is controlled according to the following equation
\begin{eqnarray}
	\mathbf \theta_{1,j}&=&((\mathbf{N}-{j})*(\mathbf{w}_{max}-\mathbf{w}_{min}))/(\mathbf{N}-{1}) + \mathbf{w}_{min}
\end{eqnarray}
where $\mathbf{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights respectively. While updating the particle, we made sure all dimensions of the solution represented by the particle were within bounds using a set of reflection boundary conditions shown below.

\begin{eqnarray}
\end{eqnarray}

\begin{algorithm}
\If{$x_{i,j}^{old} < x_{i}^{min}$}{  	
  ${x}_{i,j}^{new}={x}_{i,j}^{old}+(x_{i}^{min} - {x}_{i,j}^{old})$
  \If{${x}_{i,j}^{new}>x_{i}^{max}$}{
    ${x}_{i,j}^{new}=x_{i}^{max}$
  }
}
\If{$x_{i,j}^{old} > x_{i}^{max}$}{  	
  ${x}_{i,j}^{new}={x}_{i,j}^{old}+({x}_{i,j}^{old}-x_{i}^{max})$
  \If{${x}_{i,j}^{new}<x_{i}^{min}$}{
    ${x}_{i,j}^{new}=x_{i}^{min}$
  }
}
\end{algorithm}

After every $\mathbf{g}$  function evaluations, the particles within all sub-swarms were mixed and then randomly redistributed to a new sub-swarm. The particles were then again updated according Eq.1. This process continued till $\mathcal{FR}*\mathbf{N}$ number of functions evaluations, where $\mathcal{FR}$ represents the fraction of evaluations with the multi-swarms. The fraction of evaluations within the swam phase is based on the drop in error. If the error does not drop by 1\% for a certain number of evaluations the search within swarm phase is terminated. At the end of these function evaluations, we froze all the solutions represented by various particles and chose the particle with best solution among $\mathcal{GL}_{1} \cdots \mathcal{GL}_{NS}$ as the initial candidate vector $\mathcal{G}$ for the remaining $({1}-\mathcal{FR})*\mathbf{N}$ number of function evaluations.

This particle was then updated according to the following rule

\begin{equation}
  \mathcal{G}_{new}(\mathbf{J})=\begin{cases}
    \mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J}), & \text{if $\mathcal{G}_{new}(\mathbf{J})<\mathcal{G}(\mathbf{J})$}.\\
    \mathcal{G}(\mathbf{J}), & \text{otherwise}.
  \end{cases}
\end{equation}

where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$. $\sigma$ is the amplitude of perturbation given by following equation:
\begin{eqnarray}
	\sigma& = &\mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
\end{eqnarray}
where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension. The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.  This function is a monotonically decreasing function that decreases with the number of function evaluations. $\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function:
\begin{eqnarray}
	%\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
	\mathcal{P}_{j}&=&{1}-\log(j/ (({1}-\mathcal{FR})*\mathbf{N}))
\end{eqnarray}
%where $\beta$ is the perturbation frequency probability modulator. 
Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase. These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$. The decrease of weight function in equation 4, reflection boundary conditions in equation 5, the update function described in equations 6 and 7, the selection probability in equation 8 are the means by which DDS ideas are incorporated into DOPS.
%Equation 5 describing the reflection boundary conditions, equations 6,7 and 8 are the means by which DDS ideas are incorporated into DOPS. 
The fraction of evaluations $\mathcal{FR}$  within the swarm phase is based on a switching strategy wherein the switch from swarm phase to DDS phase happens when the error due to the best solution does not drop more than 1\% of the original error, continuously for more than a prescribed number of function evaluations. This allows the solution to quickly jump out of local optima and avoid any convergence issues that are generally associated with swarm based searches.
